Week 2 Peer Graded Assignment
=============================
## Activity Monitoring - steps per five-minute time interval
```{r loadData, echo=FALSE}
# This chunk is where we load the data
activity <- read.csv("activity.csv", header = TRUE)
```
### Total number of steps taken per day.  
  
The following is a table of the total number of steps per day:
``` {r stepsPerDay, echo=FALSE,  results = "asis", message = FALSE, warning=FALSE}
library(dplyr)
# change date from factor to date
activity <- mutate(activity, date = as.Date(as.character(date), "%Y-%m-%d"))

# group data by date
activity.date <- activity %>% group_by(date)  %>% 
    summarize(total = sum(steps, na.rm = TRUE), 
              mean.steps = mean(steps, na.rm = TRUE))

library(xtable)
# instead of one long continuous table I broke the table into three parts
# this allows for less scrolling
activity.date <- mutate(activity.date, activity.date = as.character(date)) 
myTable <- data.frame(activity.date[,c(4, 2)])
myTable1 <- myTable[1:21,]
myTable2 <- myTable[22:42,]
myTable3 <- myTable[43:61,]
newrow <- data.frame(nrow = 1, ncol = 2)
newrow[1,1] = ""
newrow[1,2] = NA
names(newrow) <- names(myTable3)
myTable3 <- rbind(myTable3, newrow)
myTable3 <- rbind(myTable3, newrow)
myTableX <- cbind(myTable1, myTable2)
myTableX <- cbind(myTableX, myTable3)
xt <- xtable(myTableX)
names(xt) <- c("Date","Total Steps","Date","Total Steps", "Date", "Total Steps")
print(xt, type = "html", include.rownames = FALSE)
```
    
  
The following is a histogram of the total number of steps taken each day:
```{r histogram, echo=FALSE}
brk <- (0:25) * 1000  
activity.date <- activity %>% group_by(date)  %>% 
    summarize(total = sum(steps, na.rm = TRUE), 
              mean.steps = mean(steps, na.rm = TRUE))
par(mfrow = c(1, 1), "mar" = c(1, 1, 1, 1))
with(activity.date, hist(total, col = "darkblue", 
                     border = "white", xlab = "Total Steps", 
                        main = "Histogram of Total Steps", breaks = brk))
StepsMean <- mean(activity.date$total)
StepsMedian <- median(activity.date$total)
```

The mean of the total number of steps per day: `r StepsMean`  
The median of the total number of steps per day: `r StepsMedian`

### Daily Activity Pattern  
The time series plot the 5-minute interval and the average number of steps taken, averaged across all days.
```{r dailyactivity, echo=FALSE,  message = FALSE, warning=FALSE}
activity.interval <- activity %>% group_by(interval)  %>% 
    summarize(total = sum(steps, na.rm = TRUE), 
              mean.steps = mean(steps, na.rm = TRUE))
plot(activity.interval$interval, activity.interval$mean.steps,xlab = "Actvity Interval", 	ylab = "Mean Number of Steps Per Interval" ,
     	type = "l", main = "Time Series Plot of Mean Steps per Time Interval")
legend("topright", lwd = 1, legend = "Mean steps per time interval")
maxInterval <-  filter(activity.interval, mean.steps == max(activity.interval$mean.steps))$interval
library(sqldf)
activity.merge <- sqldf("select a.date, a.interval, a.steps, b.[mean.steps] from activity a left outer join [activity.interval] b on a.interval = b.interval")
```
  
The five-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is `r maxInterval`


```{r MissingValues, echo=FALSE, results = "asis", message = FALSE, warning=FALSE}
Missing <- is.na(activity$steps)
```
### Imputing missing values.

The total number of missing values is `r sum(Missing)`.  
We impute the missing values by using the average number of steps for that time interval.
```{r impute, echo=FALSE}
# We impute the missing values by using the average number of steps for that time  
# interval. 
activity.interval <- activity %>% group_by(interval)  %>% 
    summarize(mean.steps = mean(steps, na.rm = TRUE))

# use SQL to merge the column with the mean values
activity.merge <- sqldf("select a.date, a.interval, a.steps, b.[mean.steps] from activity a left outer join [activity.interval] b on a.interval = b.interval")

# this short function puts the missing values in
 for(i in 1:nrow(activity.merge)) {if(is.na(activity.merge[i,3])) activity.merge[i,3] = activity.merge[i,4]}
##
## activity.merge now has the original data with the missing values filled in
activity.sum <- activity.merge %>% group_by(date)  %>% 
  summarize(total = sum(steps))
with(activity.sum, hist(total, col = "darkred", 
    border = "white", xlab = "Total Steps - missing values imputed", 
    main = "Histogram of Total Steps", breaks = brk))
```
    
### Comparing Weekday values to Weekend values.  
The following two plots compare activity in average steps per time interval for weekends and weekdays.
```{r weekend, echo=FALSE}

#add day of week to each date
activity.merge  <- mutate(activity.merge, DofW = weekdays(date, abbreviate = TRUE))

# set a logical value - true if weekday, false if weekend
activity.merge  <- mutate(activity.merge, weekday = !(DofW == "Sun" | DofW == "Sat"))

# this dataset has weekend values (not weekday)
activity.Wkend <- activity.merge[!activity.merge$weekday,]

# this dataset has weekday values
activity.Wkday <- activity.merge[activity.merge$weekday,]

activity.Wkday <- select(activity.Wkday, 2:3)  %>% group_by(interval)  %>% 
  summarize(mean.steps = mean(steps))
activity.Wkend <- select(activity.Wkend, 2:3)  %>% group_by(interval)  %>% 
  summarize(mean.steps = mean(steps))
par(mfrow = c(2, 1), "mar" = c(4, 4, 2, 1))

# plot both vertically
plot(activity.Wkday$interval, activity.Wkday$mean.steps,xlab = "Actvity Interval",
      cex.axis = 0.8,  ylab = "Number of Steps" , cex.main = 0.8,
      cex.lab = 0.8, type = "l", 
      main = "Time Series Plot of Steps per Time Interval - Weekdays")
legend("topright", lwd = 1, legend = "Steps per time interval", cex = 0.5)
plot(activity.Wkend$interval, activity.Wkend$mean.steps,xlab = "Actvity Interval", 	
     ylab = "Number of Steps", col = "red", 
     cex.main = 0.8, cex.lab = 0.8, type = "l", 
     main = "Time Series Plot of Steps per Time Interval - Weekends")
legend("topright", lwd = 1, col = "red", legend = "Seps per time interval", 
       cex = 0.5)
```
  
These two plots show that on weekends activity begins later but rises to a peak more quickly than on weekdays.  There is also more activity on weekends during the day.